---
title: "Getting started with energypal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with energypal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 96
)
```

```{r libs, message = FALSE}
library(energypal)
library(ggplot2)
library(dplyr)
```

## The problem

Energy datasets do not agree on what to call things. One says `Natural Gas`,
another `nat gas`, a third `NG`; plant-level data says `COAL1` and `COAL2`.
Published figures then colour them inconsistently, so two charts of the same
system are hard to compare.

`energypal` keeps the colours in YAML files with their provenance, and resolves
whatever labels your data happens to use onto them.

```{r first}
d <- owid_energy_mix |> filter(year == max(year))
unique(d$source)

ggplot(d, aes(country, percentage, fill = source)) +
  geom_col() +
  scale_fill_energy() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

No preprocessing, no manual colour vector, no renaming.

Seven mixes in arbitrary order are hard to compare, though. A palette carries
more than colour — `carriers` records a carbon intensity per fuel — so the
countries can be ordered by how carbon-intensive their mix is:

```{r rank}
carriers <- energypal_match(unique(d$source), warn = FALSE) |>
  select(source = original, carrier = matched)

# One row per carrier that has a figure. A name can appear twice in the table -
# Nuclear is both a group and the carrier inside it - and joining on it unfiltered
# fans the data out.
intensity <- energypal_table("carriers") |>
  filter(!is.na(carbon_intensity)) |>
  distinct(carrier = name, carbon_intensity)

rank <- d |>
  left_join(carriers, by = "source") |>
  left_join(intensity, by = "carrier") |>
  group_by(country) |>
  summarise(ci = weighted.mean(carbon_intensity, percentage, na.rm = TRUE)) |>
  arrange(desc(ci))

rank

d <- d |> mutate(country = factor(country, levels = rank$country))
```

> **That is an ordering, not an emissions estimate.** The intensities are
> indicative lifecycle medians for arranging charts; see the caveat below. Every
> figure from here on uses this ordered `d`.

## Getting a palette

`energypal()` returns a named colour vector. By default you get the main level —
the carriers themselves.

```{r palette}
energypal()
```

`n` takes the first few, as `RColorBrewer::brewer.pal()` does:

```{r n}
energypal("eia", n = 5)
```

The aggregate groups above the carriers, and the fuel subtypes below them, are
available but off by default — a plotting palette wants distinct colours, and
adjacent subtypes are deliberately near-identical.

```{r flags}
length(energypal("carriers"))
length(energypal("carriers", include_groups = TRUE, include_subtypes = TRUE))
```

Names can be canonical identifiers or display labels:

```{r labels}
energypal("carriers", n = 4, label_style = "long")
```

## What is available

```{r info}
energypal_info() |> select(name, title, type, n)
```

`energypal_show()` draws them. Given several names it draws an aligned
comparison instead, which is the quickest way to see how published schemes
differ.

```{r show, fig.height = 4.5}
energypal_show(c("carriers", "eia", "ipcc", "owid"))
```

## Matching messy labels

`energypal_colors()` maps labels to colours. `energypal_match()` shows its
working, which is what you want when a label is not colouring as expected.

```{r match}
energypal_match(c("Coal", "natural gas", "Solar PV", "COAL1", "Flux Capacitor"))
```

The `method` column names the stage that resolved each label: an exact match, a
canonicalised one (case, spacing and punctuation removed), a declared alias,
a fuzzy match, or containment. Anything unresolved comes back grey rather than
silently taking a neighbour's colour.

Aliases live in the palette file, not in the package code, so your own palette
can teach the matcher your own vocabulary. More on that below.

## Ordering

Legend and stack order carry meaning in energy charts. Palettes declare
orderings by name:

```{r orders}
energypal_orders("carriers") |> select(order, kind, n)
```

```{r ordered, fig.height = 3.6}
for (o in c("carbon_intensity", "merit_order")) {
  print(
    ggplot(d, aes(country, percentage, fill = source)) +
      geom_col() +
      scale_fill_energy(order = o) +
      labs(title = o) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

The sequence changes; the colour of each carrier does not.

> **Orderings are for presentation only.** They arrange legends and chart
> series. They are indicative, not validated figures, and must not be used in
> calculations. Every declared ordering carries a note saying so, and
> `energypal_orders()` prints it:

```{r caveat}
cat(energypal_orders("carriers")$note[1])
```

If your fill column is already a factor, its level order is respected — leave
`order` alone and set the factor levels as you normally would.

## Continuous and binned scales

Resource maps need a ramp rather than a set of categories. Those palettes are
`type = "continuous"` and carry their own break points.

```{r continuous, fig.height = 3.2}
grid <- expand.grid(x = 1:60, y = 1:24)
grid$wind <- 2 + 15 * (sin(grid$x / 9) + cos(grid$y / 5) + 2) / 4

ggplot(grid, aes(x, y, fill = wind)) +
  geom_raster() +
  scale_fill_energy_b(palette = "windatlas") +
  labs(fill = "m/s") +
  theme_minimal()
```

`scale_fill_energy_b()` takes the bins from the palette; `scale_fill_energy_c()`
gives a smooth gradient instead. Passing a continuous palette to the discrete
scale, or the reverse, is an error rather than a silent half-result.

## Several series, one carrier

Plant-level data is the common case: a fleet has three coal units, two gas
turbines and four wind farms, and colouring by carrier alone draws them all
identically. `gradient = TRUE` spreads each cluster into distinct tones of its
carrier, so the fuel is still readable at a glance but the units are told apart.

```{r gradient}
energypal_colors(c("Coal_Ratcliffe", "Coal_Drax", "Coal_Cottam",
                   "CCGT_Pembroke", "CCGT_Staythorpe", "Wind_Whitelee"),
                 gradient = TRUE, warn = FALSE)
```

Each cluster is grouped by the *carrier it resolved to*, not by its colour, so
two carriers that happen to share a hex code still get their own gradients. A
carrier with a single unit keeps the plain palette colour.

The same argument works on the scale, which is how you would normally use it.
Keeping each fuel's units adjacent is what makes the clusters legible, and
`energypal_match()` gives you the carrier to sort on. Here is a whole fleet —
forty-five units across six fuels:

```{r gradient-plot, fig.height = 3.6, fig.width = 9}
set.seed(42)
unit_group <- function(prefix, k, lo, hi) {
  data.frame(unit = sprintf("%s_%02d", prefix, seq_len(k)),
             mw = round(runif(k, lo, hi)))
}

fleet <- bind_rows(
  unit_group("Coal", 7, 400, 2000), unit_group("CCGT", 9, 300, 1400),
  unit_group("Hydro", 4, 100, 900), unit_group("Wind", 14, 50, 1300),
  unit_group("Solar", 8, 20, 400), unit_group("Nuclear", 3, 900, 3200)
)

fleet <- fleet |>
  left_join(energypal_match(fleet$unit, warn = FALSE) |>
              select(unit = original, carrier = matched),
            by = "unit") |>
  mutate(carrier = factor(carrier, levels = names(energypal()))) |>
  arrange(carrier, desc(mw)) |>
  mutate(unit = factor(unit, levels = unit))

# forty-five unit names will not fit, so label each cluster once instead
clusters <- fleet |>
  group_by(carrier) |>
  summarise(at = unit[ceiling(n() / 2)], units = n(), .groups = "drop") |>
  left_join(energypal_table("carriers") |> distinct(carrier = name, label = label_short),
            by = "carrier")

ggplot(fleet, aes(unit, mw, fill = unit)) +
  geom_col() +
  scale_fill_energy(gradient = TRUE) +
  scale_x_discrete(breaks = clusters$at,
                   labels = sprintf("%s (%d)", clusters$label, clusters$units)) +
  labs(title = "A generation fleet, shaded within each fuel",
       x = NULL, y = "MW") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.x = element_blank(), legend.position = "none")
```

Every unit gets its own colour, and the fuel is still readable across the whole
chart:

```{r gradient-count}
fleet |>
  mutate(color = energypal_colors(as.character(unit), gradient = TRUE, warn = FALSE)) |>
  group_by(carrier) |>
  summarise(units = n(), colours = n_distinct(color))
```

The shades are computed in OKLAB, a perceptually uniform space, so the step
looks the same whether the carrier is near-black coal or bright yellow nuclear —
the previous HSV implementation collapsed bright colours to a single tone.
`energypal_gradient()` does this for one colour on its own:

```{r gradient-one}
energypal_gradient("#FFA500", n = 5)     # Solar
energypal_gradient("#2C2C2C", n = 5)     # Coal
```

## Your own palette

A named vector is a palette. Save it, edit it, load it back.

```{r own}
p <- energypal_create(c(Coal = "#4E4E4E", Gas = "#2E86AB", Solar = "#F6AE2D"),
                      name = "myproject")
f <- tempfile(fileext = ".yml")
energypal_write(p, f)
cat(readLines(f), sep = "\n")
```

Entries start as bare colours and can be expanded in place with labels and
aliases, without restructuring the file:

```{r own-edit}
writeLines(c(
  "meta:",
  "  name: myproject",
  "colors:",
  '  Coal: "#4E4E4E"',
  "  Gas:",
  '    _color: "#2E86AB"',
  "    _long: Natural Gas",
  "    _aliases: [nat gas, NG, methane]"
), f)

energypal_colors(c("Coal", "nat gas", "METHANE"), file = f)
```

`energypal_register()` makes a file available by name for the rest of the
session, so it can be used anywhere a built-in name works:

```{r own-register}
energypal_register("myproject", f)
energypal("myproject")
```

## Where next

- `vignette("palettes", package = "energypal")` previews every palette with its
  sources.
- `energypal_info()` lists what is installed, including palettes you register.
- Adding a palette is one YAML file and no code.
