---
title: "The palettes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The palettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  dpi = 96
)
```

```{r libs, message = FALSE}
library(energypal)
library(ggplot2)
library(dplyr)
```

Every palette is a YAML file under `inst/extdata/palettes/`, carrying its own
colours, labels, aliases and provenance. This article previews all of them.

Nothing here is hard-coded: the sections below are generated from
`energypal_info()`, so a palette added to the store appears without this file
being edited.

## What is installed

```{r info}
info <- energypal_info()

info |>
  select(name, title, type, n, unit) |>
  knitr::kable()
```

The two kinds are handled differently throughout, so it is worth naming them
once:

```{r kinds}
discrete <- info |> filter(type == "discrete") |> pull(name)
continuous <- info |> filter(type == "continuous") |> pull(name)

discrete
continuous
```

## All at a glance

Given several names, `energypal_show()` draws them as aligned rows — one per
palette, one column per entry. Differences between published schemes are easier
to see this way than in any table.

```{r overview, fig.height = 5.5}
energypal_show(discrete)
```

Entries that only some palettes define trail off to the right: IPCC's CCS
variants, EPA's sector categories, the technologies taxonomy.

Continuous palettes have no entries to align on, so several of them are drawn
stacked instead, each with its own range and unit:

```{r overview-continuous, fig.height = 2.2}
energypal_show(continuous)
```

## Discrete palettes

A palette shows only the colours its own source publishes. Report palettes
`extends: carriers`, which gives them the taxonomy — names, labels, aliases,
hierarchy, orderings — but **not** its colours. So a carrier a source is silent
about comes back unmapped rather than quietly borrowing the default:

```{r declared-only}
setdiff(names(energypal("carriers")), names(energypal("owid")))

# OWID publishes no geothermal series, and says so
energypal_match("Geothermal", palette = "owid", warn = FALSE)

# ...while the inherited alias vocabulary still resolves
energypal_colors("nat gas", palette = "owid", warn = FALSE)
```

That is why these palettes are worth choosing between: each is its source's own
set, not the default with a few edits. Three earlier palettes (`iea`, `fossil`,
`renewable`) were dropped for failing that test — `iea` differed from `carriers`
in zero carrier-level entries, because `carriers` was seeded from it.

```{r discrete, results = "asis", fig.height = 4.6}
for (nm in discrete) {
  spec <- energypal_spec(nm)
  m <- spec$meta

  cat("\n### ", if (is.null(m$title)) nm else m$title, "  \n", sep = "")
  cat("`", nm, "`\n\n", sep = "")
  if (!is.null(m$source)) cat("**Source** ", m$source, "  \n", sep = "")
  if (!is.null(m$url))    cat("**URL** <", m$url, ">  \n", sep = "")
  if (!is.null(m$extends)) cat("**Extends** `", m$extends, "`  \n", sep = "")
  if (!is.null(m$license)) cat("\n> **Licence** ", m$license, "\n", sep = "")
  if (!is.null(m$attribution)) {
    cat("\n> **Attribution** ", m$attribution, "\n", sep = "")
  }
  cat("\n")

  energypal_show(nm, label_style = "default")
  cat("\n\n")
}
```

## Continuous palettes

Resource maps need a ramp rather than a set of categories. These palettes are an
ordered list of stops and declare their own break points and unit, so a binned
scale needs no configuration.

```{r continuous, results = "asis", fig.height = 2.4}
for (nm in continuous) {
  spec <- energypal_spec(nm)
  m <- spec$meta
  brk <- as.numeric(unlist(spec$breaks))

  cat("\n### ", if (is.null(m$title)) nm else m$title, "  \n", sep = "")
  cat("`", nm, "` — family `", m$family, "`, measure `", m$measure,
      "`, unit `", m$unit, "`  \n", sep = "")
  cat(length(energypal(nm)), " stops, ", length(brk), " breaks from ",
      min(brk), " to ", max(brk), "  \n", sep = "")
  if (!is.null(m$source)) cat("**Source** ", m$source, "  \n", sep = "")
  if (!is.null(m$license)) cat("\n> **Licence** ", m$license, "\n", sep = "")
  if (!is.null(m$attribution)) {
    cat("\n> **Attribution** ", m$attribution, "\n", sep = "")
  }
  cat("\n")

  energypal_show(nm)
  cat("\n\n")
}
```

An atlas publishes several quantities and styles each differently, so each is a
separate palette named `family_measure`. They are not interchangeable. One
member of each family is its default, so the bare family name resolves to it:

```{r family}
identical(energypal("windatlas"), energypal("windatlas_speed"))
energypal_info(family = "windatlas") |> select(name, measure, default, unit)
```

Several measures published by the atlases are **not** included — wind power
density and IEC capacity-factor classes, solar DNI, DIF and GTI — because no
verified colour source for them was available, and guessing would defeat the
purpose of a store whose value is its provenance. They are better added as one
reviewable file each than approximated in bulk.

## Orderings

Order carries meaning in energy charts, so palettes declare it rather than
leaving it to whatever sequence the file happens to use.

```{r orders}
energypal_orders("carriers") |>
  select(order, kind, n, source) |>
  knitr::kable()
```

The same data under three of them. The sequence changes; the colour of each
carrier does not.

```{r ordering-demo, fig.height = 3.4}
# countries ordered most carbon-intensive first, using the palette's own
# carbon-intensity property - a presentation ordering, not an emissions estimate
latest <- owid_energy_mix |> filter(year == max(year))

# one row per carrier that has a figure: a name can appear twice in the table
# (Nuclear is both a group and the carrier inside it), which would fan the join out
intensity <- energypal_table("carriers") |>
  filter(!is.na(carbon_intensity)) |>
  distinct(carrier = name, carbon_intensity)

rank <- latest |>
  left_join(energypal_match(unique(latest$source), warn = FALSE) |>
              select(source = original, carrier = matched), by = "source") |>
  left_join(intensity, by = "carrier") |>
  group_by(country) |>
  summarise(ci = weighted.mean(carbon_intensity, percentage, na.rm = TRUE)) |>
  arrange(desc(ci))

d <- latest |> mutate(country = factor(country, levels = rank$country))
for (o in c("carbon_intensity", "merit_order", "renewability")) {
  print(
    ggplot(d, aes(country, percentage, fill = source)) +
      geom_col() +
      scale_fill_energy(order = o) +
      labs(title = o, x = NULL, y = "%") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

An ordering declared at carrier level covers the subtypes beneath it: an entry
the sequence does not name sorts with its nearest named ancestor, so each fuel's
family stays together.

```{r subtypes}
energypal_table("carriers", order = "carbon_intensity") |>
  select(name, type) |>
  head(10)
```

> **Orderings, and the figures behind them, are for presentation only.** They
> arrange legends and chart series. They are indicative values, not validated
> figures, and must not be used in calculations. Every declared ordering and
> numeric property carries a note saying so, which travels with the data:

```{r caveat}
tab <- energypal_table("carriers")
attr(tab$carbon_intensity, "note")
attr(tab$carbon_intensity, "source")
```

## Shades within a carrier

Plant-level data repeats a carrier many times over — a fleet has several coal
units and a dozen wind farms. `energypal_gradient()` spreads one palette colour
into distinct tones so those clusters stay apart without losing the fuel.

```{r shades, fig.height = 4.6, fig.width = 8}
base <- energypal("carriers", n = 10)
n_sh <- 5

op <- par(mai = c(0.2, 1.15, 0.4, 0.2))
plot.new()
plot.window(xlim = c(0.5, n_sh + 0.5), ylim = c(length(base) + 0.5, 0.5))
for (i in seq_along(base)) {
  sh <- energypal_gradient(base[[i]], n = n_sh)
  for (j in seq_len(n_sh)) {
    rect(j - 0.5, i - 0.42, j + 0.5, i + 0.42, col = sh[j], border = "white", lwd = 0.6)
  }
  mtext(names(base)[i], side = 2, at = i, las = 1, line = 0.3, cex = 0.75)
}
title(main = sprintf("%d shades of each carrier", n_sh), cex.main = 0.95)
par(op)
```

The middle column is the palette colour; the gradient spreads symmetrically
around it in OKLAB rather than only lightening, which is what keeps a bright
carrier like Solar from collapsing to one tone.

`energypal_colors(gradient = TRUE)` applies this to data, grouping by the
carrier each label *resolved to*:

```{r shades-data}
energypal_colors(c("Coal_1", "Coal_2", "Coal_3", "Wind_N", "Wind_S", "Solar_1"),
                 gradient = TRUE, warn = FALSE)
```

Grouping by resolved carrier rather than by colour matters where two entries
share a hex code — `FossilCoal` and `OtherFossilCoal` are both `#2C2C2C`, and
grouping on the colour would merge them into one gradient.

## The same chart, four ways

```{r comparison, fig.height = 3.4}
for (p in c("carriers", "eia", "ipcc", "owid")) {
  print(
    ggplot(d, aes(country, percentage, fill = source)) +
      geom_col() +
      scale_fill_energy(palette = p) +
      labs(title = p, x = NULL, y = "%") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

## Adding a palette

One YAML file in `inst/extdata/palettes/`, or anywhere on disk if you pass
`file =`. No code changes, no registration in a list, no test edits — the
validator tests iterate over the directory, so a new file is picked up and
checked automatically.

```yaml
meta:
  name: myreport
  title: "Colours for the 2026 report"
  source: "Figure 4.2"
  license: "Our own colours"   # required: how they were obtained
  extends: carriers            # optional: inherit labels and aliases
colors:
  FossilCoal: "#3A3A3A"        # state only what this palette publishes
  FossilGas: "#5599CC"
```

`extends` applies each entry wherever that name occurs in the parent, at any
depth, keeping the labels and aliases it already had — and *only* those two
entries get a colour. A palette that wants to be standalone simply omits it.

```{r validate}
f <- tempfile(fileext = ".yml")
writeLines(c("meta:", "  name: myreport", "  license: Our own colours",
             "  extends: carriers",
             "colors:", '  FossilCoal: "#3A3A3A"'), f)

energypal_validate(energypal_spec(file = f))
energypal_colors(c("Coal", "coal power"), file = f)   # aliases inherited
energypal_colors("Solar", file = f, warn = FALSE)     # not declared here
```

### Provenance is required

`energypal_validate()` refuses a palette that does not say where its colours
came from, and one declaring a CC BY licence without the citation that licence
requires. That rule exists because the store is only worth having if every entry
can be traced.

Colours reach a palette one of two ways, and `meta.license` always says which:
reproduced from a downloadable Work published under a licence that permits it —
the atlas poster maps — or approximated by eye from a published figure, where the
source publishes no colour specification at all. The second is the common case,
which is why those palettes are titled *(approximation)*.

Your own palettes are exempt. `energypal_register()` validates leniently, and
`energypal_validate(..., strict = FALSE)` checks structure alone:

```{r lenient}
mine <- energypal_create(c(Coal = "#FF0000", Solar = "#00FF00"), name = "mine")
energypal_validate(mine, strict = FALSE)
```
