---
title: "Getting Started with energypal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with energypal}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5, 
  eval = FALSE
)
```

```{r setup}
library(energypal)
library(ggplot2)
```

## Introduction

The `energypal` package provides color palettes and functions specifically designed for energy-related data visualizations. This vignette demonstrates the main features and shows how to create effective energy visualizations.

## Available Palettes

First, let's explore the available energy color palettes:

```{r palette-overview}
# View all available palette names
get_energy_palette_names()

# Get a specific palette
iea_palette <- get_energy_palette("iea_primary")
print(iea_palette)
```

## Displaying Palettes

You can visualize any palette to see the colors and their mappings:

```{r display-palette, fig.width=7, fig.height=3}
# Display the IEA primary palette
display_energy_palette("iea_primary")
```

## Smart Color Mapping

One of the key features of `energypal` is intelligent mapping of energy source names to appropriate colors:

```{r smart-mapping}
# Exact matching
sources <- c("Coal", "Natural Gas", "Solar", "Wind")
colors <- map_energy_colors(sources)
print(colors)

# Fuzzy matching for messy data
messy_sources <- c("coal power", "nat gas", "solar pv", "wind energy")
fuzzy_colors <- map_energy_colors(messy_sources, method = "fuzzy")
print(fuzzy_colors)
```

## Using with ggplot2

The package integrates seamlessly with ggplot2 through custom scale functions and convenience geoms. We now rely on the real-world curated `owid_energy_mix` dataset (which replaced the former simulated example).

```{r ggplot-basic, fig.width=8, fig.height=5}
# Load OWID subset
data(owid_energy_mix)

# Focus on United States for clarity
usa_mix <- subset(owid_energy_mix, country == "United States")

# Basic energy mix chart (automatic ordering + legend limited to present sources)
# Pre-order the data to ensure consistent legend and stack ordering
usa_mix_ordered <- usa_mix
# canonical_order <- get_canonical_order("iea_primary", "carbon_intensity")
present_sources <- unique(usa_mix$source)
canonical_present <- intersect(canonical_order, present_sources)
non_canonical <- setdiff(present_sources, canonical_present)
# Build the same order as the scale with direction = -1
final_order <- c(non_canonical, rev(canonical_present))
usa_mix_ordered$source <- factor(usa_mix_ordered$source, levels = final_order)

ggplot(usa_mix_ordered, aes(x = year, y = generation_twh, fill = source)) +
  geom_col(position = "stack") +  # Use regular geom_col with pre-ordered data
  scale_fill_energy(direction = -1, data = usa_mix) +
  labs(
    title = "United States Electricity Generation Mix",
    x = "",
    y = "TWh",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Stacked Area Chart (Shares to TWh)

```{r stacked-area, fig.width=8, fig.height=5}
ggplot(usa_mix, aes(x = year, y = generation_twh, fill = source)) +
  geom_area() +
  scale_fill_energy(palette = "renewable_focus", direction = 1) +
  labs(
    title = "United States Electricity Generation (TWh)",
    x = "Year",
    y = "Generation (TWh)",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Energy Generation by Country

```{r generation-chart, fig.width=8, fig.height=6}
# Load real-world energy generation data
data(owid_energy_mix)

# Focus on recent year and selected countries for clear visualization
latest_year <- max(owid_energy_mix$year)
selected_countries <- c("United States", "China", "Germany", "Brazil", "France")
generation_data <- subset(owid_energy_mix, year == latest_year & 
                         country %in% selected_countries)

ggplot(generation_data, aes(x = country, y = generation_twh, fill = source)) +
  geom_col() +
  scale_fill_energy(palette = "technology_focus") +
  coord_flip() +
  labs(
    title = "Electricity Generation by Energy Source and Country",
    x = "Country",
    y = "Generation (TWh)",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Working with Different Palettes

Different palettes are suitable for different types of visualizations:

```{r compare-palettes, fig.width=10, fig.height=8}
library(gridExtra)
p1 <- ggplot(usa_mix, aes(year, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1) +
  scale_fill_energy(palette = "iea_primary") +
  ggtitle("IEA Primary") + theme_minimal() + theme(legend.position = "none")
p2 <- ggplot(usa_mix, aes(year, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1) +
  scale_fill_energy(palette = "epa_ghg") +
  ggtitle("EPA GHG") + theme_minimal() + theme(legend.position = "none")
p3 <- ggplot(usa_mix, aes(year, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1) +
  scale_fill_energy(palette = "renewable_focus") +
  ggtitle("Renewable Focus") + theme_minimal() + theme(legend.position = "none")
# grid.arrange(p1, p2, p3, ncol = 3)
```

## Creating Custom Palettes

You can also create and add your own custom palettes:

```{r custom-palette}
# Define a custom palette
my_custom_palette <- c(
  "Fossil Fuels" = "#8B4513",
  "Renewables" = "#228B22", 
  "Nuclear" = "#FFD700",
  "Storage" = "#9370DB"
)

# Validate the palette
validate_energy_palette(my_custom_palette)

# Add to the collection (for the session)
add_energy_palette("my_custom", my_custom_palette)
```

## Utility Functions

The package includes several utility functions for working with energy colors:

```{r utilities}
# Create a comprehensive mapping dictionary
mapping_dict <- create_energy_mapping(include_variations = TRUE)
cat("Number of mappings:", length(mapping_dict), "\n")
cat("First 10 mappings:\n")
print(head(mapping_dict, 10))

# Interpolate colors between energy sources
coal_to_solar <- interpolate_energy_colors("Coal", "Solar", n = 5)
print(coal_to_solar)
```

## Direction vs. Palette Direction vs. Stack Direction

You can now independently control:

1. `direction` – legend/source ordering (reverses factor/limits)
2. `palette_direction` – reverses color sequence without changing labels
3. `stack_direction` – physical stacking (handled by `geom_energy_col()` / `geom_energy_bar()` or `position_energy_stack()`)

```{r direction-palette-stack, fig.width=11, fig.height=6}
latest <- max(usa_mix$year)
usa_latest <- subset(usa_mix, year == latest)

# Base
g_base <- ggplot(usa_latest, aes(country, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1) +
  scale_fill_energy(direction = 1) + ggtitle("Base (direction=1, stack=-1)") + theme_minimal() + theme(legend.position="none")

# Reverse legend order only
g_dir <- ggplot(usa_latest, aes(country, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1, direction = -1) +
  scale_fill_energy(direction = -1) + ggtitle("Legend Reversed (direction=-1)") + theme_minimal() + theme(legend.position="none")

# Reverse palette colors only
g_pal <- ggplot(usa_latest, aes(country, percentage, fill = source)) +
  geom_energy_col(stack_direction = -1, palette_direction = -1) +
  scale_fill_energy(palette_direction = -1) + ggtitle("Palette Reversed (palette_direction=-1)") + theme_minimal() + theme(legend.position="none")

# Reverse physical stack only
g_stack <- ggplot(usa_latest, aes(country, percentage, fill = source)) +
  geom_energy_col(stack_direction = 1) +
  scale_fill_energy(direction = 1) + ggtitle("Stack Inverted (stack_direction=1)") + theme_minimal() + theme(legend.position="none")

gridExtra::grid.arrange(g_base, g_dir, g_pal, g_stack, nrow = 2)
```

## Color Variations for Duplicate Sources

When you have multiple instances of the same energy source (e.g., multiple coal plants, different wind farms), you can use color variations to distinguish them:

```{r color-variations, fig.width=8, fig.height=3}
# Show how variations work
display_energy_palette("iea_primary", show_variations = TRUE, n_variations = 5, 
                      show_labels = TRUE)
```

The variations create ordered progressions (from subtle to more pronounced) and use smart labeling:
- For ≤3 variations: Individual labels ("Coal (1)", "Coal (2)", "Coal (3)")  
- For >3 variations: Single centered range label ("Coal (1-5)", "Nuclear (1-10)")

This avoids label clutter while clearly showing the variation scope.

When using `energypal` for energy visualizations:

1. **Choose appropriate palettes**: Use `renewable_focus` for renewable energy analysis, `fossil_focus` for fossil fuel analysis, etc.

2. **Use intelligent mapping**: Let the package map your energy source names automatically using `map_energy_colors()`

3. **Validate custom palettes**: Always validate custom palettes with `validate_energy_palette()`

4. **Consider accessibility**: The palettes are designed with reasonable contrast, but always consider your audience

5. **Document your choices**: Include palette information in your analysis documentation

## Conclusion

The `energypal` package provides a comprehensive toolkit for creating consistent, professional energy visualizations. By using standardized color palettes inspired by major energy organizations, your visualizations will be both aesthetically pleasing and familiar to energy sector audiences.
